---
title: "Introduction-to-datimutils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction-to-datimutils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(httptest)
start_vignette("Introduction-to-datimutils")
```

```{r, echo=FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(purl = NOT_CRAN)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datimutils)
```
# Logging In
Logging in can be done with the `loginToDATIM` function. This function can either take the name of a keyring stored on the OS or a datim config file with credentails stored as plain text. In the case of a config file that is nested with multiple credentials, by default the function will use the one titled "dhis" but this can be changed with the `config_path_level` argument.  
```{r, eval = NOT_CRAN}
#logging in generates a cookie so the other funcitons can be used freely without credentials
loginToDATIM(ring = "DatimLogin")
```
In the case that a keyring needs to be created it can be done with the `makeKeyring` function. This function will first ask for the **KEYRING password and then the SECRET** the dialog box that asks for said credentials however, will not look any different, so make sure to keep this in mind.
```c
makeKeyring(ring ="burner_key", service ="burner_url", username = "burner_name") 
```



```{r, include=FALSE}
end_vignette()
```
