---
title: "Introduction-to-datimutils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction-to-datimutils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = T
)
```

```{r setup}
library(datimutils)
```

Before querying the DATIM api, you must log into a DATIM instance. There are three basic ways to do so within datimutils. All three utilize the `loginToDATIM` function.

By default the `loginToDATIM` function will create an object containing details about the session. This object is named *d2_default_session*, and it is created in the environment from which the `loginToDATIM` function is called. In this vignette we are assuming you are logging in from the global environment. All datimutils functions will search for the d2_default_session by default, allowing you to make multiple DATIM api requests through datimutils without individually specifying unique login objects.

> Note: By default, this session will expire after 60 minutes. When it does, you will need to re-authenticate 


  
The most basic method is to pass your username and password directly into the `loginToDATIM` function.
  
```{r eval = F}
loginToDATIM(
  base_url = 'play.dhis2.org/2.36/',
  username = 'admin',
  password = 'district'
)
```

The other two methods require the use of a configuration file. Make sure that you remember the location of this config file, as you will need it to pass into the `loginToDATIM` function.

In this case, our config file is a JSON file called, *play.json*.

There are two ways to login with this config file. 

In the first, you input your baseurl, username, and password.

```{r eval = F}
{
"dhis": {
  "baseurl": "play.dhis2.org/2.36/",
  "username": "admin",
  "password": "district"
}
}
```
Alternatively, if the the entry for password is blank such that the file looks like this
```{r eval = F}
{
"dhis": {
  "baseurl": "play.dhis2.org/2.36/",
  "username": "admin",
  "password": ""
}
}
```
The function will look in your keyring for the password, and if not found, will prompt you to enter a password and will store it automatically under that username and service (baseurl) for the next login. A user may prefer storing their passwords in their keyring because it allows them to avoid storing passwords in plain text on their computer, which is a potential security risk.   
>Note: keyrings operate differently on different operating systems


In both methods, call the `loginToDATIM` function as so:
```{r}
# logging in generates a cookie so the other funcitons can be used freely without credentials
loginToDATIM(config_path = "play.json")
```

There is a wrapper for every metadata endpoint within datimutils.
Here are calls to  dataElements and categoryOptionGroups for example:
```{r}
getDataElements(
    "FTRrcoaog83")
getCatOptionGroups(
    "OK2Nr4wdfrZ")
```

Here is one to the Indicators endpoint:
```{r}
  getIndicators(
        "ReUHfIn0pTQ")
```

These helpers will create repeated values and sorting based on input
```{r}
# make a random order and duplicated uid group
 groups <- rep(c(
    "gzcv65VyaGq", "uYxK4wmcPqA", "RXL3lPSK8oG"), 20)
 rows <- sample(length(groups))
 groups <- c("gzcv65VyaGq", "uYxK4wmcPqA", "RXL3lPSK8oG",
              groups[rows])

  data <-
    getOrgUnitGroups(
      groups,
      fields = "code,name,id")

# even though the actual api call only returns the 3 unique uids in a random
# order, the wrapper returns it in the order called.
str(data)
```

The `getMetadata` function can be used to call metadata returning endpoints. It should be used for more complicated metadata calls with more than one filter. It takes an endpoint, any number of filters, and fields as its basic arguments. For example, to call the organisationUnits endpoint with two filters the following call can be used:

```{r}
data <- getMetadata(
  end_point = "organisationUnits",
  "organisationUnitGroups.name:eq:District",
  "children.id:in:[YuQRtpLP10I,fwH9ipvXde9]",
  fields = "id,name,level"
)

head(data)
```

This is equivalent to:

```{r}
# notice the use of non-standard evaluation
data <- getMetadata(
  end_point = organisationUnits,
  organisationUnitGroups.name %.eq% "District",
  children.id %.in% c("YuQRtpLP10I","fwH9ipvXde9"),
  fields = "id,name,level"
)

head(data)
```

Data is usually returned in its most base class:
```{r}
# returns a vector
data <- getMetadata(
  end_point = "organisationUnitGroups",
  fields = "name"
)
print(data)
```

many different types of calls call be made, here we have two filters passed in as a vector. This call returns a nested data frame
```{r}
data <- getMetadata(
  end_point = "organisationUnits",
  c(
    "name:like:Baoma",
    "level:eq:3"
  ),
  fields = ":all"
)
str(data$translations)
```

There is a filter format helper for every comparison operator in Datim, here like is used.
```{r}
data <- getMetadata(
  end_point = "organisationUnits",
  name %.Like% "Sierra Leone"
)
```

Here is a list of the api comparison operators and their datimutils equivalents:

```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
| DHIS2 Operator| infix operator  | Description                                     |
|---------------|:---------------:|------------------------------------------------:|
| eq            | %.eq%           | Equality                                        |
| !eq           | %.~eq%          | Inequality                                      |
| like          | %.Like%         | Case sensitive string, match anywhere           |
| !like         | %.~Like%        | Case sensitive string, not match anywhere       |
| \\$like       | %.^Like%       | Case sensitive string, match start             |
| !\\$like      | %.~^Like%      | Case sensitive string, not match start         |
| like\\$       | %.Like$%       | Case sensitive string, match end               |
| !like\\$      | %.~Like$%      | Case sensitive string, not match end           |
| ilike         | %.like%         | Case insensitive string, match anywhere         |
| !ilike        | %.~like%        | Case insensitive string, not match anywhere     |
| \\$ilike      | %.^like%       | Case insensitive string, match start           |
| !\\$ilike     | %.~^like%      | Case insensitive string, not match start       |
| ilike\\$      | %.like$%       | Case insensitive string, match end             |
| !ilike\\$     | %.~like$%      | Case insensitive string, not match end         |
| gt            | %.gt%           | Greater than                                    |
| ge            | %.ge%           | Greater than or equal                           |
| lt            | %.lt%           | Less than                                       |
| le            | %.le%           | Less than or equal                              |
| token         | %.token%        | Match on multiple tokens in search property     |
| !token        | %.~token%       | Not match on multiple tokens in search property |
| in            | %.in%           | Find objects matching 1 or more values          |
| !in           | %.~in%          | Find objects not matching 1 or more values      |
"
cat(tabl)
```

There is a analytics endpoint helper, `getAnalytics`. It takes arguments for dimensions and filters, and can also take any number of additional specifications as the first argument, which will get pasted on the end of the api call made. Here are some examples:

```{r}
#Get ANC: Key Coverages
#this call uses only dimension arguments, dx, ou, and pe
data <- getAnalytics("displayProperty=NAME",
dx = c("Uvn6LCg7dVU","OdiHJayrsKo","sB79w2hiLp8"), pe = c("THIS_YEAR"),
ou = c("ImspTQPwCqd"))

head(data)

#Get Infrastructure: Public facilities 2014 and 2013
#this call uses a filter ou argument, which are followed by _f
data <- getAnalytics("displayProperty=NAME",
dx = c("v27b2y9kKh7","vEBapQFZ1AA","DwEVHQvMnDZ"), pe = c("THIS_YEAR"),
ou_f = c("ImspTQPwCqd"))

head(data)
```

To use `getAnalytics` you can pass in the uids into their respective arguments and will be formatted correctly, but there are also two operators which can format strings for use in analytics api calls.

```{r}
a %.d% c("123","456")
a %.f% c("123","456")
```
